---
title: "Week 10b: Activity Key"
format: pdf
editor: source
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(sp)
library(knitr)
library(gstat)
```

### Last Time

- Fitting GP models
- Kriging with plug-in estimates

### This week

- Stationarity
- Locations close are more similar, but how? -> Variograms

---

## Stationarity


Assume the spatial process, $\boldsymbol{Y}(\boldsymbol{s})$ has a mean, ${\mu}(\boldsymbol{s})$, and that the variance of $\boldsymbol{Y}(\boldsymbol{s})$ exists everywhere.

The process is __strictly stationary__ if: for any $n \geq 1$, any set of $n$ sites $\{\boldsymbol{s}_1, \dots, \boldsymbol{s}_n\}$ and any $\boldsymbol{h} \in \mathcal{R}^r$ (typically r =2), the distribution of $(\boldsymbol{Y}(\boldsymbol{s}_1). \dots, \boldsymbol{Y}(\boldsymbol{s}_n))$ is the same as $(\boldsymbol{Y}(\boldsymbol{s_1+h}) \dots, \boldsymbol{Y}(\boldsymbol{s_n + h})$


Weak stationarity, or second-order stationarity of a spatial process, requires a constant mean and covariance that is a function of $\boldsymbol{h}$, $Cov(\boldsymbol{Y}(\boldsymbol{s}),\boldsymbol{Y}(\boldsymbol{s +h})) = C(\boldsymbol{h}),$ where $C(\boldsymbol{h})$ is a covariance function that only requires the separation vector $\boldsymbol{h}.$

Typically with spatial models, the process is assumed to be mean zero as covariates explain the mean structure. So second-order stationarity is primarily focused on the covariance structure.


A third kind of stationarity, known as intrinsic stationarity, describes the behavior of differences in the spatial process, rather than the data.

Intrinsic stationarity assumes $E[\boldsymbol{Y}(\boldsymbol{s + h}) -\boldsymbol{Y}(\boldsymbol{s}) ] =0$ then define
$$E[\boldsymbol{Y}(\boldsymbol{s + h}) - \boldsymbol{Y}(\boldsymbol{s})]^2 = Var(\boldsymbol{Y}(\boldsymbol{s + h}) - \boldsymbol{Y}(\boldsymbol{s})) = 2 \gamma(\boldsymbol{h}),$$
which only works, and satisfies intrinsic stationarity, if the equation only depends on $\boldsymbol{h}$.

## Variograms


Variograms, defined as $2\gamma( \boldsymbol{h})$ are often used to visualize spatial patterns:

\vfill
- If the distance between points is small, the variogram is expected to be small
    
\vfill
- As the distance between points increases, the variogram increases

\vfill

There is a mathematical link between the covariance function $C(\boldsymbol{h})$ and the variogram $2\gamma(\boldsymbol{h})$.


## Linear semivariogram
$$\gamma(d)=
\begin{cases}
\tau^2 + \sigma^2d \; \; \text{if } d > 0\\
0 \; \; \text{otherwise}
\end{cases}$$

```{r}
tau.sq <- 1
sigma.sq <- 1
d <- seq(0,3, by =.01)
lin.gam <- tau.sq + sigma.sq * d
lin.var <- data.frame(d=d, lin.gam = lin.gam )
ggplot(data = lin.var, aes(x=d, y=lin.gam)) + 
  geom_line() + 
  ylim(0,4) + 
  ylab('linear variogram') + 
  theme_bw()
```



### Nugget, sill, partial-sill, and range
Nugget is defined as $\gamma(d): \; d \rightarrow 0_+$

\vfill

Sill is defined as $\gamma(d): \; d \rightarrow \infty$

\vfill

Partial sill is defined as the $\text{sill} - \text{nugget}$

\vfill

Range is defined as the first point where $\gamma(d)$ reaches the sill. Range is sometimes (1/$\phi$) and sometimes $\phi$ depending on the paremeterization.


## Spherical semivariogram: 
$$\gamma(d)=
\begin{cases}
\tau^2 + \sigma^2 \; \; \text{if } d \geq 1/ \phi\\
\tau^2 + \sigma^2\left[\frac{3\phi d}{2} - \frac{1}{2} (\phi d)^3 \right]\\
0 \; \; \text{otherwise}
\end{cases}$$

- Sketch, or generate in R, a spherical semivariogram

- On this figure label the nugget, sill, partial sill, and range.


## Spherical semivariogram: Solution

```{r}
#| echo: FALSE
tau.sq <- 1
sigma.sq <- 2
range <- 2
phi <- 1/range
d <- seq(0,100, by =.01)
sph.gam <- tau.sq + sigma.sq * (3 * phi * d / 2 - .5 * (phi*d)^3)
sph.gam[d >= range] <- tau.sq + sigma.sq
sph.var <- data.frame(d=d, sph.gam = sph.gam )
ggplot(data = sph.var, aes(x=d, y=sph.gam)) + 
  geom_line() + ylim(0,4) + 
  ylab('spherical variogram') + 
  annotate('text', x=0.1, y = 0.9, label = 'nugget') + 
  annotate('text', x=2.5, y = 3.1, label = 'sill') + 
  annotate('text', x=2.2, y = .1, label = 'range', color = 'blue') +  
  annotate("segment", x = 2, xend = 2, y = 0, yend = 4, colour = "blue", linetype =3, size = .5) +  
  annotate("segment", x = 2, xend = 3, y = 1, yend = 1,colour = "red", linetype =3, size = .5) +  
  annotate("segment", x = 2, xend = 3, y = 3, yend = 3,colour = "red", linetype =3, size = .5) +  
  annotate("segment", x = 2.5, xend = 2.5, y = 1, yend = 3,colour = "red", linetype =3, size = .5) + 
  annotate('text', x=2.2, y = 2, label = 'partial sill', color = 'red') 
```


## Exponential 

We saw the exponential covariance earlier in class, what is the mathematical form of the covariance?

\vfill

$$C(d)=
\begin{cases}
\tau^2 + \sigma^2 \; \; \text{if } d = 0\\
\sigma^2 \exp(- d/ \phi) \; \; \text{if } d > 0
\end{cases}$$

\vfill

The variogram is 
$$\gamma(d)=
\begin{cases}
\tau^2 + \sigma^2(1 - \exp(- d / \phi)) \; \; \text{if } d > 0\\
0 \; \; \text{otherwise} 
\end{cases}$$


```{r}
#| echo: false
exp.gam <- tau.sq + sigma.sq * (1 - exp( - phi * d))
exp.var <- data.frame(d=d, exp.gam = exp.gam )
 ggplot(data = exp.var, aes(x=d, y=exp.gam)) + geom_line() + ylim(0,4) + 
   ylab('exponential variogram') + 
   annotate('text', x=0.1, y = 0.9, label = 'nugget') + 
   annotate('text', x=2.9, y = 2.65, label = 'sill?') + 
   annotate('text', x=2.2, y = .1, label = 'range?', color = 'blue') +  
   annotate("segment", x = 2, xend = 2, y = 0, yend = 4,
   colour = "blue", linetype =3, size = .5) + theme_bw()
```


The sill is only reached asymptotically when $$\lim_{d\rightarrow \infty} exp(- d / \phi) \rightarrow 0$$

\vfill

Thus, the range, defined as distance where semivariogram reaches the sill is $\infty$

\vfill

The *effective range* is defined as the distance where there is *effectively* no spatial structure. Generally this is determined when the correlation is .05. 

\vfill

\begin{eqnarray*}
\exp(-d_o / \phi) &=&.05 \\
d_0 / \phi &=& - \log(.05) \\
d_0 &\approx& 3 \phi
\end{eqnarray*}

```{r}
#| echo: false
d <- seq(0,8, by = .01)
exp.gam <- tau.sq + sigma.sq * (1 - exp( - phi * d))
exp.var <- data.frame(d=d, exp.gam = exp.gam )
ggplot(data = exp.var, aes(x=d, y=exp.gam)) + geom_line() + ylim(0,4) + ylab('exponential variogram') + annotate('text', x=0.1, y = 0.9, label = 'nugget') + annotate('text', x=7.9, y = 3.1, label = 'sill') + annotate('text', x=6.6, y = .1, label = 'effective range', color = 'blue') +  annotate("segment", x = 6, xend = 6, y = 0, yend = 4,
  colour = "blue", linetype =3, size = .5) + theme_bw()
```

\vfill

## More Semivariograms: Equations
Gaussian: 
$$\gamma(d)=
\begin{cases}
\tau^2 + \sigma^2(1 - \exp(-\phi^2 d^2)) \; \; \text{if } d > 0\\
0 \; \; \text{otherwise} 
\end{cases}$$

\vfill

Powered Exponential:
$$\gamma(d)=
\begin{cases}
\tau^2 + \sigma^2(1 - \exp(-|\phi d|^p)) \; \; \text{if } d > 0\\
0 \; \; \text{otherwise} 
\end{cases}$$

\vfill

Matern$\acute{\text{e}}$rn:
$$\gamma(d)=
\begin{cases}
\tau^2 + \sigma^2\left[1 - \frac{(2\sqrt{\nu}d \phi)^\nu}{2^{\nu-1}\Gamma(\nu)} K_\nu (2 \sqrt{\nu} d \phi)\right] \; \; \text{if } d > 0\\
0 \; \; \text{otherwise} 
\end{cases},$$
where $K_\nu$ is a modified Bessel function and $\Gamma()$ is a Gamma function.

\vfill



## #Variogram Creation: How?

```{r, echo=T,}
data(meuse)
meuse.small <- meuse %>% select(x, y, copper) %>% as_tibble()
meuse.small %>% head(15) %>% kable()

meuse |>
  ggplot(aes(x = x , y=y, size = copper, color = copper)) +
  geom_point() +
  scale_color_viridis_b() +
  theme_minimal()

```

### Variogram Creation: Steps

1. Calculate distances between sampling locations
2. Choose grid for distance calculations
3. Calculate empirical semivariogram
$$\hat{\gamma}(d_k) = \frac{1}{2N(d_k)} \sum_{\boldsymbol{s}_i,\boldsymbol{s}_i, \in N(d_k)}\left[ \boldsymbol{Y}(\boldsymbol{s}_i) - \boldsymbol{Y}(\boldsymbol{s}_j)  \right]^2,$$
where $$N(d_k) = \{(\boldsymbol{s}_i,\boldsymbol{s}_j): ||\boldsymbol{s}_i -\boldsymbol{s}_j|| \in I_k\}$$ and $I_k$ is the $k^{th}$ interval.
4. Plot the semivariogram



### Variogram Creation: How - Step 1

Calculate Distances between sampling locations

```{r, echo= T}
#dist(meuse.small)
dist.mat <- dist(meuse.small %>% select(x,y))
```

### Variogram Creation: How - Step 2

Choose grid for distance calculations

```{r, echo= t}
cutoff <- max(dist.mat) / 3 # default maximum distance
num.bins <-  15
bin.width <- cutoff / 15
```

### Variogram Creation: How - Step 3

Calculate empirical semivariogram

```{r}
dist.sq <- dist(meuse.small$copper)^2

vario.dat <- data.frame(dist = as.numeric(dist.mat), 
                        diff = as.numeric(dist.sq)) %>% 
  mutate(bin = floor(dist / bin.width) + 1) %>% 
  filter(bin < 16) %>% 
  group_by(bin) %>% 
  summarize(emp.sv = .5 * mean(diff)) 
vario.dat

```

## Variogram Creation: How - Step 4

Plot empirical semivariogram
```{r}
vario.dat %>% 
  ggplot(aes(x=bin, y= emp.sv)) + 
  geom_point() + xlim(0,15) + ylim(0,800) + 
  scale_x_continuous(breaks=c(5, 10, 15), labels =c('500','1000','1500')) + 
  ggtitle('Empirical Semivariogram for Copper in Meuse Data Set') + 
  xlab('distance (m)') + ylab('Empirical Semivariogram')
```

Now given this empirical semivariogram, how to we choose a semivariogram (and associated covariance structure) and estimate the parameters in that function??


## Variogram Creation: R function

```{r, eval = T}
coordinates(meuse) = ~x+y
variogram(copper~1, meuse) %>% plot()
```